Kernel Square-Loss Exemplar Machines

### 2016-11-15 sec_eval_raf.tex
version J-1

\subsection{Datasets and evaluation protocol} \label{eval:protocol}
We perform our experiments on three standard datasets for image retrieval. 
\begin{itemize}
    \item The INRIA \emph{Holidays} dataset \cite{holidays} consists of 1491 images divided in 500 groups of matching images. We manually rotate by 90 degrees some images that are not in their natural orientation to compensate for the fact that CNN features are not rotation invariant~\cite{babenko14,Arandjelovic15,RaToCh16}
    \item The \emph{Oxford5k} dataset \cite{oxford}, which consists of 5063 images separated in 55 groups of matching images, each group associated to a landmark of Oxford. We use the "full" crop, ignoring the region of interest of each image.
    \item The \emph{Oxford105k} dataset \cite{oxford}, large-scale dataset containing the same images and queries from Oxford5k plus \emph{Flickr100k}, a collection of $10^5$ distractor images Flickr.
\end{itemize}
As pool of negative images to build SLEM, we use the Flickr100k for both Holidays and Oxford5k. When training on Oxford105k, we us instead the \emph{Paris} dataset \cite{PhiChIsSiZi08} as negative samples.
%In both datasets, each group contains one query image, the other images in the group being the only correct answers to the query.
%For each query image, we calculate its similarity to all other images in the database and rank them in decreasing order.
%The average precision of a group is derived from the ranking of the images of the group for the similarity with the corresponding query image.
%The final mean average precision (mAP) for a dataset is the mean of the average precision over all its groups.
%As pool of negative images to build SLEM, we use the Flickr100k collection \cite{oxford}, composed of $10^5$ images from Flickr. For a full rank decomposition, we use only a subset of it, containing between 6000 and 15000 images.
%As stated in section \ref{low-rank} and further discussed in section \ref{time-scale}, a full rank decomposition does not scale well for bigger number of negative samples. For low rank decomposition, we use all 100000 images.

At evaluation time, for a dataset that consists of $p$ images and $q$ query images, we calculate the $p\times q$ \emph{similarity matrix} $S$, where each of its $q$ columns is the matching scores of the query image with all the $p$ images. \hlc{RR: maybe take out this paragraph completely.}

\begin{table*}[t!]
\begin{center}
\setlength{\tabcolsep}{.2em}
%\begin{tabular}{|c|c|c|c|c|c|c|c|}
%\small
\begin{tabular}{c@{\hskip 1em}cccc@{\hskip 1em}cccc@{\hskip 1em}cc}%@{\hskip 1em}c}
\toprule
Dataset & \multicolumn{4}{c}{\textbf{Holidays}} & \multicolumn{4}{c}{\textbf{Oxford 5k}} & \multicolumn{2}{c}{\textbf{Oxford 105k}} \\%& \textbf{Oxford 105k}\\
\midrule
Model, features & VLAD  & SPoC & AlexNet & NetVLAD & VLAD & SPoC & AlexNet & NetVLAD & SPoC & NetVLAD\\% & SPoC\\
\midrule
Baseline            & 72.7 & 76.5 & 68.2  &  85.4 & 46.3 & 54.4 & 40.6 & 67.5 & 50.1 & - \\% & \\
%Whitening           & -    & -    & -    & -    & -\\
PCAW                & 75.5 & 81.7 & 69.2 & 88.3 & 50.9  & 63.7 & 45.0 & 69.1 & 55.5 & - \\% & \\
LDA                 & 54.7 & 82.2 & 64.1 & 74.3 & 29.6  & 62.2 & 42.5 & 72.7 & 52.4 & - \\
ESVM~\cite{ZePe15}  & $77.5^3$ & $84.0^3$ & 71.3 & $91.4^2$ & 57.2  & 62.1 & 43.9 & 72.5 & 56.5 & - \\% & \\ 
Linear SLEM         & $78.0^2$ & 82.3 & 72.1 & $91.3^3$ & \ul{59.3}  & 64.1 & 46.2 & 72.9 & 56.7 & - \\% & \\ /42.2
Gaussian SLEM (16)  & 76.8 & 80.3 & 71.2 & $91.4^2$ & 52.8 & 63.0 & 43.5 & 71.9 & 55.8 & - \\
Gaussian SLEM (32)  & 77.4 & 81.7 & $72.0^3$ & $91.4^2$ & 54.9 & 63.1 & 44.0 & 71.1 & 56.0 & - \\
Gaussian SLEM (fr)  & \bf{78.1} & $86.2^2$ & \bf{72.9} & \bf{91.7} & 59.0   & \ul{64.9} & 47.0 & \ul{74.4} & 59.5 & - \\% & \\ 
Poly SLEM (16)      & 76.9 & 82.3 & 71.4 & $91.3^3$ & 53.0 & 63.6 & 43.6 & 71.4 & 56.1 & - \\
Poly SLEM (32)      & 77.3 & 82.4 & $72.1^2$ & \bf{91.7} & 54.9 & 63.6 & 44.1 & 71.6 & 56.3 & - \\
Poly SLEM   (fr)    & \bf{78.1} & \bf{86.3}  & \bf{72.9} & \bf{91.7} & \ul{59.3}  & 64.8 & \ul{47.3} & \ul{74.1} & 62.5 & - \\
\hline
\end{tabular}
\caption{Mean average precision results for INRIA Holidays and Oxford buildings datasets, expressed as percentages. In this table, we present our results for VLAD \cite{Delhumeau2013}, sum-pooling of convolutional features (SPoC) \cite{babenko15}, activation coefficients from the previous-to-last CNN layer (AlexNet) \cite{Krizhevsky2012} and activation of NetVLAD layer~\cite{Arandjelovic15}. In parenthesis, the rank of he decomposition ('fr' for full rank decomposition)}
\end{center}
\label{fullrank}
\end{table*}

\subsection{Which kernel to choose?}
For the experiment of SLEM as a feature encoder, we tested two different kernels -- Gaussian and polynomial -- each with a scalar parameter $\gamma$:
\begin{align}
    k_1(x,y) = e^{-\gamma\|x-y \|^2}; \ 
    k_2(x,y) = x^Ty+\gamma(x^Ty)^2. \label{kernels}
\end{align}
\textbf{Gaussian SLEM.} The radial basis function kernel $k_1$ is a
well known reproducing kernel, used for classification with support vector machines.\\
\textbf{Poly SLEM.} The polynomial kernel $k_2$ is a reproducing kernel often used in natural language processing.\\
%Note the non-kernel version of SLEM is equivalent to a SLEM for the linear kernel. In order to distinguish it from other versions, we will refer to the non-kernel version as \textbf{Linear SLEM} for the remaining of this paper.
%\textbf{SPM SLEM} The spatial pyramid matching kernel of $L+1$ levels in Equation \ref{k:spm} take as input a set of local descriptors and its location in pyramidal bins \cite{spk}.

\subsection{Base visual features}
We test our feature encoder for four different base features; the hand-crafted VLAD image representation and three learned features derived from the activation coefficients of deep Convolutional Neural Networks.

@@

The first CNN-based features we use consist of the activation coefficients of the previous-to-last layer of the AlexNet architecture \cite{Krizhevsky2012}, based on a publicly available pre-trained model \cite{jia2014caffe}. These are also the features used in \cite{ZePe15}.

@@

\subsection{Image retrieval results}
We use the base features of the previous subsection as baseline. Since \textit{Babenko et al.}~\cite{babenko15} and \textit{Arandjelovi\'c et al.}~\cite{Arandjelovic15} have improved retrieval results by applying PCA followed by whitening to their features, we also apply this post-processing to our base features as second baseline (PCAW), compressing its its dimension to half of the original. We then compare the baselines with the original ESVM, LDA and several variants of our approach (SLEM), since all the those methods are based on similar \hlc{??? (I can't read the last word you wrote.} The results are presented in Table \ref{fullrank}.

Linear SLEM performs similarly to ESVM despite being much more time efficient (see discussion in section \ref{time-scale}). 
The fact that a hinge-loss classifier does not outperform a square-loss classifier can seem counter-intuitive, but both were shown to be equivalent for binary classification under mild constraints~\cite{YeXi07}.

We perform both Gaussian SLEM and Polynomial SLEM with two decompositions: one full-rank CCD decomposition signaled by (fr) and two low-rank KPCA decompositions signaled by the rank of the decomposition. We train our exemplar classifiers for between $6000$ and $15000$ negative samplers. For all the experiments we calibrate the regularization cost $\lambda$, as well as the parameter $\gamma$.

The full-rank variant outperform all methods for all base features, although the gains when compared to linear SLEM are not always significant (\emph{e.g.} for VLAD features). It is interesting to notice that LDA, ESVM and Linear SLEM do not seem to outperform one another or the baselines consistently for all features and datasets.
%For VLAD and AlexNet, we considere the version presented in~\cite{ZePe15} as \textit{Baseline} and a rotated, whitened and compressed to half its original dimensions as \textit{PCA+whitening}. 
%As for the version of each feature to which we apply the other algorithms, it changes according to the dataset. 
%For SPoC, we apply LDA, ESVM and SLEM to the baseline, as a replacement of PCA. 
%For NetVLAD, we use the rotated and compressed version of the trained networks correspondent to each dataset as suggested by~\cite{Arandjelovic15}. 
%For Holidays we obtain better results with whitened features and for Oxford5k, non-whitened.

@@

\begin{table}[t]
\begin{center}
\setlength{\tabcolsep}{.2em}
\small
\begin{tabular}{l|c|ll}
\toprule
Features & dim & \textbf{Hol.} & \textbf{Ox5k} \\%& \textbf{Oxford 105k}\\
\midrule
Babenko \textit{et al.}\cite{babenko15}  & 256 & 80.2 & 58.9 \\ %& 57.8\\
Radenovi\'c \textit{et al.} \cite{RaToCh16}   & 256 & 81.5 & \un{77.4} \\
Arandjelovi\'c \textit{et al.} \cite{Arandjelovic15}& 256 & 86.0 & 62.5 \\ %& - \\
Kalantidis  \textit{et al.} \cite{KaMeOs16}   & 256 & 83.1 & 65.4 \\
SPoC + Linear SLEM & 256 & 81.5 & 64.7 \\
SPoC + Poly SLEM & 288 & 80.1 & 63.6 \\
SPoC + Poly SLEM & 320 & 81.8 & 63.6 \\
NetVLAD + Linear SLEM & 256 & \un{88.5} & 65.9 \\
NetVLAD + Poly SLEM & 288 & 87.7 & 65.5 \\
NetVLAD + Poly SLEM & 320 & 88.3 & 65.6 \\
\midrule
Radenovi\'c \textit{et al.} \cite{RaToCh16}   & 512 & 82.5 & 79.7 \\
Arandjelovi\'c \textit{et al.} \cite{Arandjelovic15}& 512 & 86.7 & 65.6 \\
Kalantidis \textit{et al.} \cite{KaMeOs16}   & 512 & 84.9 & 68.2 \\
Gordo \textit{et al.} \cite{GoAlReLa16} & 512 & 89.1^{\dag} & \bf{83.1}^{\dag} \\
SPoC + Linear SLEM & 512 & 82.3 & 64.1 \\
SPoC + Poly SLEM & 544 & 82.3 & 63.0 \\
SPoC + Poly SLEM & 576 & 82.4 & 63.1 \\
NetVLAD + Linear SLEM & 512 & 89.3 & 72.3 \\
NetVLAD + Poly SLEM & 544 & \un{89.9} & 71.9 \\
NetVLAD + Poly SLEM & 576 & \un{89.9} & 72.3 \\
\midrule
Arandjelovi\'c \textit{et al.} \cite{Arandjelovic15}& 4096 & 88.3 & 69.1 \\
%Li \textit{et al.} \cite{LiLaHa15}& - & 89.2^{\dag} & 73.7 \\
NetVLAD + Linear SLEM & 4096 & 91.3 & 72.9 \\
NetVLAD + Poly SLEM & 4128 & 91.3 & 71.2 \\
NetVLAD + Poly SLEM & 4160 & \bf{91.7} & 71.7 \\
%Zepeda \textit{et al.} \cite{ZePe15} & \underline{78.3} & - & 71.8 & - & 57.5 & - & 44.6 & - & - \\
%Babenko \textit{et al.} \cite{babenko15} & - & 81.8 & - & - & - & \textbf{65.7} & - & - & 57.8\\
\hline 
\end{tabular}
\caption{Compared results to state-of-the-art features at similar dimensions, without re-ranking or query augmentation. The results using Poly SLEM or Gaussian SLEM add 32 or 64 dimensions to the original feature (for $r=16$ or $r=32$, respectively). Underlined results are the best at each dimension bracket and bold results are the general best. $\dag$ indicates the previous state-of-the-art.}
\label{sota}
\end{center}
\end{table}

>>>>

\subsection{Datasets and evaluation protocol} \label{eval:protocol}
We perform our experiments on three standard datasets for image retrieval:
\begin{itemize}
    \item The INRIA \emph{Holidays} dataset \cite{holidays}, which consists of 1491 images divided in 500 groups of matching images. We manually rotate by 90 degrees some images that are not in their natural orientation to compensate for the fact that CNN features are not rotation invariant~\cite{Arandjelovic15,babenko14,RaToCh16}
    \item The \emph{Oxford5k} dataset \cite{oxford}, which consists of 5063 images separated in 55 groups of matching images, each group associated to a landmark of Oxford. We use the ``full'' crop, ignoring the region of interest of each image.
    \item The \emph{Oxford105k} dataset \cite{oxford}, a large-scale dataset containing the same images and queries from Oxford5k plus \emph{Flickr100k}, a collection of $10^5$ distractor Flickr images.
\end{itemize}
As pool of negative images to build SLEM, we use the Flickr100k for both Holidays and Oxford5k. When training on Oxford105k, we us instead the \emph{Paris} dataset \cite{PhiChIsSiZi08} as negative samples.
%In both datasets, each group contains one query image, the other images in the group being the only correct answers to the query.
%For each query image, we calculate its similarity to all other images in the database and rank them in decreasing order.
%The average precision of a group is derived from the ranking of the images of the group for the similarity with the corresponding query image.
%The final mean average precision (mAP) for a dataset is the mean of the average precision over all its groups.
%As pool of negative images to build SLEM, we use the Flickr100k collection \cite{oxford}, composed of $10^5$ images from Flickr. For a full rank decomposition, we use only a subset of it, containing between 6000 and 15000 images.
%As stated in section \ref{low-rank} and further discussed in section \ref{time-scale}, a full rank decomposition does not scale well for bigger number of negative samples. For low rank decomposition, we use all 100000 images.
At evaluation time, for a dataset that consists of $p$ images and $q$ query images, we calculate the $p\times q$ \emph{similarity matrix} $S$, where each of its $q$ columns is the matching scores of the query image with all the $p$ images.

\begin{table*}[t!]
\begin{center}
\setlength{\tabcolsep}{.2em}
%\begin{tabular}{|c|c|c|c|c|c|c|c|}
%\small
\begin{tabular}{c@{\hskip 1em}cccc@{\hskip 1em}cccc@{\hskip 1em}cc}%@{\hskip 1em}c}
\toprule
Dataset & \multicolumn{4}{c}{\textbf{Holidays}} & \multicolumn{4}{c}{\textbf{Oxford 5k}} & \multicolumn{2}{c}{\textbf{Oxford 105k}} \\%& \textbf{Oxford 105k}\\
\midrule
Model, features & VLAD  & SPoC & AlexNet & NetVLAD & VLAD & SPoC & AlexNet & NetVLAD & SPoC & NetVLAD\\% & SPoC\\
\midrule
Baseline            & 72.7 & 76.5 & 68.2  &  85.4 & 46.3 & 54.4 & 40.6 & 67.5 & 50.1 & - \\% & \\
PCAW                & 75.5 & 81.7 & 69.2 & 88.3 & 50.9  & 63.7 & 45.0 & 69.1 & 55.5 & - \\% & \\
LDA                 & 54.7 & 82.2 & 64.1 & 74.3 & 29.6  & 62.2 & 42.5 & 72.7 & 52.4 & - \\
ESVM~\cite{ZePe15}  & 77.5^3 & 84.0^3 & 71.3 & 91.4^2 & 57.2^3 & 62.1 & 43.9 & 72.5 & 56.5 & - \\% & \\ 
Linear SLEM         & 78.0^2 & 82.3 & 72.1 & 91.3^3 & \bf{59.3}  & 64.1^3 & 46.2^3 & 72.9^3 & 56.7^3 & - \\% & \\ /42.2
Gaussian SLEM (16)  & 76.8 & 80.3 & 71.2 & 91.4^2 & 52.8 & 63.0 & 43.5 & 71.9 & 55.8 & - \\
Gaussian SLEM (32)  & 77.4 & 81.7 & 72.0^3 & 91.4^2 & 54.9 & 63.1 & 44.0 & 71.1 & 56.0 & - \\
Gaussian SLEM (fr)  & \bf{78.1} & 86.2^2 & \bf{72.9} & \bf{91.7} & 59.0^2 & \bf{64.9} & 47.0^2 & \bf{74.4} & 59.5^2 & - \\% & \\ 
Poly SLEM (16)      & 76.9 & 82.3 & 71.4 & 91.3^3 & 53.0 & 63.6 & 43.6 & 71.4 & 56.1 & - \\
Poly SLEM (32)      & 77.3 & 82.4 & 72.1^2 & \bf{91.7} & 54.9 & 63.6 & 44.1 & 71.6 & 56.3 & - \\
Poly SLEM   (fr)    & \bf{78.1} & \bf{86.3}  & \bf{72.9} & \bf{91.7} & \bf{59.3}  & 64.8^2 & \bf{47.3} & 74.1^2 & \bf{62.5} & - \\
\hline
\end{tabular}
\caption{Mean average precision results for INRIA Holidays and Oxford buildings datasets, expressed as percentages. In this table, we present our results for VLAD \cite{Delhumeau2013}, sum-pooling of convolutional features (SPoC) \cite{babenko15}, activation coefficients from the previous-to-last CNN layer (AlexNet) \cite{Krizhevsky2012} and activation of NetVLAD layer~\cite{Arandjelovic15}. In parenthesis, the rank of he decomposition ('fr' for full rank decomposition). For each column, we \textbf{bold} the best results and index the second and third best.}
\label{fullrank}
\end{center}
\end{table*}

\subsection{Which kernel to choose?}
We have tested two different kernels, each with a scalar parameter $\gamma$:
%For the experiment of SLEM as a feature encoder, we tested two different kernels -- Gaussian and polynomial -- each with a scalar parameter $\gamma$:
\begin{align}
    k_1(x,y) = e^{-\gamma\|x-y \|^2}; \ 
    k_2(x,y) = x^Ty+\gamma(x^Ty)^2. \label{kernels}
\end{align}
\textbf{Gaussian SLEM.} The radial basis function kernel $k_1$ is a
well known reproducing kernel, used for classification with support vector machines.\\
\begin{align}
    k_1(x,y) = e^{-\gamma\|x-y \|^2};
\end{align}
\textbf{Poly SLEM.} The polynomial kernel $k_2$ is a reproducing kernel often used in natural language processing.\\
\begin{align}
    k_2(x,y) = x^Ty+\gamma(x^Ty)^2.
\end{align}
%Note the non-kernel version of SLEM is equivalent to a SLEM for the linear kernel. In order to distinguish it from other versions, we will refer to the non-kernel version as \textbf{Linear SLEM} for the remaining of this paper.
%\textbf{SPM SLEM} The spatial pyramid matching kernel of $L+1$ levels in Equation \ref{k:spm} take as input a set of local descriptors and its location in pyramidal bins \cite{spk}.

\subsection{Base visual features}
We test our feature encoder for four different base features: the hand-crafted VLAD image representation and three learned features derived from the activation coefficients of deep Convolutional Neural Networks.

@@

The first CNN features we use consist of the activation coefficients of the previous-to-last layer of the AlexNet architecture \cite{Krizhevsky2012}, based on a publicly available pre-trained model \cite{jia2014caffe}. These are also the features used in \cite{ZePe15}.

@@

\subsection{Image retrieval results}
We use the base features of the previous subsection as baseline. Since Babenko and Lemptisky~\cite{babenko15} and Arandjelovi\'c \etal~\cite{Arandjelovic15} have improved retrieval results by applying PCA followed by whitening to their features, we also apply this post-processing to our base features as a second baseline (PCAW), compressing base feature dimension to half of the original. We then compare the baselines with the original ESVM, LDA and several variants of our approach (SLEM), since all the those methods are based on similar ideas. The results are presented in Table \ref{fullrank}. For the large-scale dataset that is Oxford105k, we limit our experiments to our best performing base features, SPoC and NetVLAD.

Linear SLEM performs similarly to ESVM despite being much more time efficient (Fig. \ref{fullrank:results}). 
The fact that a hinge-loss classifier does not outperform a square-loss classifier can seem counter-intuitive, but both have been shown to be equivalent for binary classification under mild constraints~\cite{YeXi07}.

We use both Gaussian SLEM and Polynomial SLEM with two decompositions: one full-rank CCD decomposition indicated by (fr) and two low-rank KPCA decompositions indicated by the rank of the decomposition. We train our exemplar classifiers for between $6000$ and $15000$ negative samplers. For all the experiments we calibrate the regularization cost $\lambda$, as well as the parameter $\gamma$ similarly to the calibration in~\cite{ZePe15}.

The full-rank variant outperform all methods for all base features, although the gains when compared to linear SLEM are not always significant (\emph{e.g.} for VLAD features). We notice significant improvement for SPoC in both Holidays and Oxford and for AlexNet and NetVLAD in Oxford. It is interesting to notice that LDA, ESVM and Linear SLEM do not seem to outperform one another or the baselines consistently for all features and datasets.
%For VLAD and AlexNet, we considere the version presented in~\cite{ZePe15} as \textit{Baseline} and a rotated, whitened and compressed to half its original dimensions as \textit{PCA+whitening}. 
%As for the version of each feature to which we apply the other algorithms, it changes according to the dataset. 
%For SPoC, we apply LDA, ESVM and SLEM to the baseline, as a replacement of PCA. 
%For NetVLAD, we use the rotated and compressed version of the trained networks correspondent to each dataset as suggested by~\cite{Arandjelovic15}. 
%For Holidays we obtain better results with whitened features and for Oxford5k, non-whitened.

@@

\begin{table}[t]
\begin{center}
\setlength{\tabcolsep}{.2em}
\small
\begin{tabular}{l|c|c|ll}
\toprule
Features & rank & dim & \textbf{Hol.} & \textbf{Ox5k} \\%& \textbf{Oxford 105k}\\
\midrule
Babenko \textit{et al.}\cite{babenko15}         & - & 256 & 80.2 & 58.9 \\ 
Radenovi\'c \textit{et al.} \cite{RaToCh16} & - & 256 & 81.5 & \un{77.4} \\
Arandjelovi\'c \textit{et al.} \cite{Arandjelovic15} & - & 256 & 86.0 & 62.5 \\ 
Kalantidis  \textit{et al.} \cite{KaMeOs16}   & - & 256 & 83.1 & 65.4 \\
SPoC + Linear SLEM & - & 256 & 81.5 & 64.7 \\
SPoC + Poly SLEM & 16 & 288 & 80.1 & 63.6 \\
SPoC + Poly SLEM & 32 & 320 & 81.8 & 63.6 \\
NetVLAD + Linear SLEM & - & 256 & \un{88.5} & 65.9 \\
NetVLAD + Poly SLEM & 16 & 288 & 87.7 & 65.5 \\
NetVLAD + Poly SLEM & 32 & 320 & 88.3 & 65.6 \\
\midrule
Radenovi\'c \textit{et al.} \cite{RaToCh16}   & - & 512 & 82.5 & 79.7 \\
Arandjelovi\'c \textit{et al.} \cite{Arandjelovic15}& - & 512 & 86.7 & 65.6 \\
Kalantidis \textit{et al.} \cite{KaMeOs16}   & - & 512 & 84.9 & 68.2 \\
Gordo \textit{et al.} \cite{GoAlReLa16} & - & 512 & 89.1^{\dag} & \bf{83.1}^{\dag} \\
SPoC + Linear SLEM & - & 512 & 82.3 & 64.1 \\
SPoC + Poly SLEM   & 16 & 544 & 82.3 & 63.0 \\
SPoC + Poly SLEM   & 32 & 576 & 82.4 & 63.1 \\
NetVLAD + Linear SLEM & - & 512 & 89.3 & 72.3 \\
NetVLAD + Poly SLEM & 16 & 544 & \un{89.9} & 71.9 \\
NetVLAD + Poly SLEM & 32 & 576 & \un{89.9} & 72.3 \\
\midrule
Arandjelovi\'c \textit{et al.} \cite{Arandjelovic15}& - & 4096 & 88.3 & 69.1 \\
%Li \textit{et al.} \cite{LiLaHa15}& - & 89.2^{\dag} & 73.7 \\
NetVLAD + Linear SLEM & - & 4096 & 91.3 & 72.9 \\
NetVLAD + Poly SLEM & 16 & 4128 & 91.3 & 71.2 \\
NetVLAD + Poly SLEM & 32 & 4160 & \bf{91.7} & 71.7 \\
%Zepeda \textit{et al.} \cite{ZePe15} & \underline{78.3} & - & 71.8 & - & 57.5 & - & 44.6 & - & - \\
%Babenko \textit{et al.} \cite{babenko15} & - & 81.8 & - & - & - & \textbf{65.7} & - & - & 57.8\\
\hline 
\end{tabular}
\caption{Compared results to state-of-the-art features at similar dimensions, without re-ranking or query augmentation. The results using Poly SLEM or Gaussian SLEM add 32 or 64 dimensions to the original feature (for $r=16$ or $r=32$, respectively). Underlined results are the best at each dimension bracket and bold results are the general best. $\dag$ indicates the previous state-of-the-art.}
\label{sota}
\end{center}
\end{table}

### 2016-11-15 sec_intro.tex
version J-1

An effective image 
representation is a crucial component of image retrieval system
where, given some query image,  images must be retrieved from a large 
database, then ranked according to their degree of similarity with the 
query.
%Robust image representations are a crucial component of a vast number of computer vision applications. 
%Within these, the \emph{image retrieval} application is an important example wherein a query image is provided to the system, which must in turn find all matching images from within a large, unannotated database. 
The matching process is done entirely based on the pixel content of the query and database images, and the search must be robust to large image variations due to camera pose, color differences and scene illumination, amongst others.

The advances in image representation enabled the success of new systems in this challenging task. %An image representation function commonly maps a given input image to a vector in a fixed dimensional space called the image \emph{feature vector}. 
Indeed, many successful feature representations for image retrieval rely on unsupervised models such as $K$-means \cite{Delhumeau2013} or Gaussian mixture models \cite{Perronnin2010} and, before the neural networks \textit{renaissance}, these representations would outperform methods that exploit supervised learning of image features directly~\cite{Bilen2015,Rana}.
Today, with the popularization of convolutional architectures, global image descriptors are often obtained by aggregation and/or pooling the last convolutional layers \cite{babenko15,GoAlReLa16,KaMeOs16} or addition of new differentiable layers to an existing architecture~\cite{Arandjelovic15}.
%One of the main reasons for this is the lack of adequately large and varied supervised datasets that are expensive to collect.
%A now pervasive example of an image feature representation is that consisting of the activation coefficients extracted from the previous-to-last layer of Convolutional Neural Networks (CNNs) \cite{Krizhevsky2012}. Although CNNs are trained in a fully-supervised manner for the image classification task, they have been shown to transfer well not only to new, unseen classes \cite{Oquaba,Chatfield2014}, but also to alternate tasks such as object detection \cite{Girshick2014} and, importantly, image retrieval \cite{Sharif}.

The exemplar support vector machine (ESVM), originally proposed by Malisiewicz {\it et al.} \cite{Malisiewicza}, leverages the availability of large, unannotated pools of images within the context of supervised learning. It uses a large generic pool of images as a set of negative examples, while using a single image (the \emph{exemplar}) as a positive example. Given these training sets, a classifier is learned that can  generalize well, despite the drastically limited size of the set of positive examples.

Zepeda \emph{et al.} \cite{ZePe15} propose to instead treat the weights of the resulting classifier as a new feature vector for image retrieval. %This new ESVM feature is derived from \emph{base} feature representations ({\it e.g.,} CNN activation features) of the exemplar and the negative pool.
%An extension \cite{ZePe15} of the ESVM formulation discussed above instead treats the resulting  classifier as an \emph{enhanced} representation of the image feature representing the exemplar image (the \emph{base} feature representation).
An ESVM feature is extracted from each database image, as well as from the query image, by treating each image as an exemplar while keeping a fixed pool of generic negative images. Searching amounts to computing distances between the query and the database ESVM features. Note that ESVM features can be derived from arbitrary \emph{base} features ({\it e.g.}, CNN activations) of the exemplar and the images in the generic negative pool. %An interesting interpretation of this approach is that the ESVM features represent what is unique about the exemplar relative to the pool of generic negatives, which is the role of a classifier.

One important drawback of the ESVM feature encoding approach is that computing the  classifier requires solving an optimization problem for each positive example. This can be time consuming for the large negative pool sizes required for good ESVM feature performance. In this work we propose using the square loss instead of the hinge loss, in effect converting the ESVM problem into a ridge regression that can be solved in closed form. The parameters of the corresponding classifier, which we call the squared loss exemplar machine (SLEM), had been shown to both generalize LDA \cite{Koba15} and, for the general case of multiple positive samples (LS-SVM)~\cite{lssvm}, form features with comparable binary classification performance while drastically reducing the computational cost.

Since computing the SLEM features requires inverting a large matrix related to the training set's covariance matrix, we propose an efficient way to compute this inverse that exploits the fact that only a single (positive) example changes in the training set when computing SLEM features for different images. We show experimentally that our representation matches and even improves upon the performance of ESVM features on two standard datasets using a wide range of base features.

We also introduce a kernelized variant of SLEM that enjoys similar computational advantages but improves retrieval performances. Further computational efficiency is obtained using a kernel principal components analysis (KPCA) decomposition of the negative samples.
%low rank incomplete Cholesky decomposition of the negative pool kernel matrix.

>>>>

This paper addresses the problem of designing an image representation suitable for content-based retrieval, in particular supporting effective (discriminative) and
efficient (fast) comparisons between a query picture and images stored in some large database.
%Robust image representations are a crucial component of a vast number of computer vision applications. 
%Within these, the \emph{image retrieval} application is an important example wherein a query image is provided to the system, which must in turn find all matching images from within a large, unannotated database. 
These comparisons are done entirely based on the pixel content of the query and database images, and the search must be robust to large image variations due to camera pose, color differences and scene illumination, amongst others.

%The advances in image representation enabled the success of new systems in this challenging task. %An image representation function commonly maps a given input image to a vector in a fixed dimensional space called the image \emph{feature vector}. 
Many successful image representations for retrieval rely on unsupervised models such as $K$-means \cite{Delhumeau2013} or Gaussian mixture models \cite{Perronnin2010,PeSan10} and, before the neural networks \textit{renaissance}, these representations would outperform methods that exploit supervised learning of image features directly~\cite{Bilen2015,Rana}.
Today, with the success of convolutional architectures, global image descriptors are often obtained by aggregating and/or pooling the last convolutional layers \cite{babenko15,GoAlReLa16,KaMeOs16} or by addition of new differentiable layers to an existing architecture~\cite{Arandjelovic15}.
%One of the main reasons for this is the lack of adequately large and varied supervised datasets that are expensive to collect.
%A now pervasive example of an image feature representation is that consisting of the activation coefficients extracted from the previous-to-last layer of Convolutional Neural Networks (CNNs) \cite{Krizhevsky2012}. Although CNNs are trained in a fully-supervised manner for the image classification task, they have been shown to transfer well not only to new, unseen classes \cite{Oquaba,Chatfield2014}, but also to alternate tasks such as object detection \cite{Girshick2014} and, importantly, image retrieval \cite{Sharif}.

The exemplar support vector machine (ESVM), originally proposed by Malisiewicz {\it et al.} \cite{Malisiewicza}, leverages the availability of large, unannotated pools of images within the context of supervised learning. It uses a large generic pool of images as a set of negative examples, while using a single image (the \emph{exemplar}) as a positive example. Given these training sets, a SVM classifier is learned that can generalize well, despite the drastically limited size of the set of positive examples.

Zepeda and P\'erez \cite{ZePe15} propose to treat instead the weights of the resulting classifier as a new feature vector for image retrieval. %This new ESVM feature is derived from \emph{base} feature representations ({\it e.g.,} CNN activation features) of the exemplar and the negative pool.
%An extension \cite{ZePe15} of the ESVM formulation discussed above instead treats the resulting  classifier as an \emph{enhanced} representation of the image feature representing the exemplar image (the \emph{base} feature representation).
An ESVM feature is extracted from each database and query image, by treating it as the only positive sample (an \emph{exemplar}) while keeping a fixed pool of generic negative images. Searching amounts to computing distances between the query and the database ESVM features. Note that ESVM features can be derived from arbitrary \emph{base} features ({\it e.g.}, CNN activations) of the exemplar and the images in the generic negative pool. %An interesting interpretation of this approach is that the ESVM features represent what is unique about the exemplar relative to the pool of generic negatives, which is the role of a classifier.

One drawback of the ESVM feature encoding approach is that computing the  classifier requires solving an optimization problem for each positive example (\ie, each query and each database image). This can be time consuming for the large negative pool sizes required for good ESVM feature performance. In this work, we propose using the square-loss instead of the hinge loss, in effect converting the ESVM problem into a ridge regression that can be solved in closed form. 
We dub the corresponding classifier a {\em square-loss exemplar machine} (or {\em SLEM}). The square-loss has been used before to replace the hinge loss in classification tasks (e.g.,~\cite{lssvm,YeXi07}), and the SLEM has also been used to compare ESVMs to classical classifiers such as the linear discriminant analysis (or {\em LDA})~\cite{Koba15}. We focus here on the role of SLEMs as feature encoder for image retrieval instead of categorization.
%The parameters of the corresponding classifier, which we call the squared loss exemplar machine (SLEM), had been shown to both generalize LDA \cite{Koba15} and, for the general case of multiple positive samples (LS-SVM)~\cite{lssvm}, form features with comparable binary classification performance while drastically reducing the computational cost.

Since computing the SLEM features requires inverting a large matrix related to the training set's covariance matrix, we propose an efficient way to compute this inverse that exploits the fact that only a single (positive) example changes in the training set when computing SLEM features for different images. We show experimentally that our representation matches and even improves upon the performance of ESVM features on three standard datasets using a wide range of base features.

We also introduce a kernelized variant of SLEM that enjoys similar computational advantages and improves retrieval performances. 
This kernel approach allows the use of any global image feature (\eg HoG~\cite{hog}, BoW~\cite{BoW04}, spatial pyramids~\cite{spk}) as base representation, independent of its use on retrieval.
Further computational efficiency is obtained using low-rank factorization methods, such as the incomplete Cholesky decomposition (ICD) or kernel principal components analysis (KPCA), to decompose the kernel matrix of negative samples.
%low rank incomplete Cholesky decomposition of the negative pool kernel matrix.
